Arterial pulsations are not powerful enough to drive perivascular drainage of solutes from the brain

### 2017-01-11 paper.tex
added doctoral prize funding information
moved step function to back of section

The BM consists of a complex matrix of proteins (extracellular matrix, ECM) that can be interpreted as a porous medium through which soluble metabolites drain via the ISF. We model flow through this medium, in the standard fashion, by Darcy's law in a 2D cylindrical coordinate system. Darcy's law states that the flux $\q$ is related to the fluid pressure gradient $\boldsymbol{\nabla} p$ by
\beq\label{eq:darcy}
  \q = -K(p_z) \boldsymbol{\nabla} p
\eeq
where $K(p_z)$\footnote{Note that the permeability $K$ used here is the ratio of the intrinsic permeability $k$ of the ECM to the viscosity $\mu$ of ISF, that is $K = k / \mu$} is the pressure gradient dependent permeability of the ECM. We allow the permeability $K$ to be a function of the axial component of the ISF pressure gradient $\partial p(z,t) / \partial z = p_z(z,t)$. In order to obtain a net flux up the vessel, counter to the blood flow, we require this function to be increasing in $p_z$, representing the notion that it is harder to push a flow in the positive $z$-direction ($p_z<0$ and $K(p_z)$ small) than in the negative $z$-direction ($p_z>0$ and $K(p_z)$ large)
\begin{equation}
  K(p_z) = \begin{cases}
    K_0 & p_z > 0\\
    K_1 & p_z \leq 0.
  \end{cases}
\end{equation}

@@

Because the BM thickness is much smaller than the vessel length it is possible, on using the lubrication approximation, to derive a 1D model (in $z$) for the BM thickness $h(z,t)$; this takes the form
\beq
\frac{\dd}{\dd t} \left( R_i(z,t) \cdot h(z,t) \right) = \frac{\dd}{\dd z} \left( R_i(z,t) \cdot h(z,t) \cdot K(p_{z}) \cdot p_{z} \right),\label{eq:pvs_model}
\eeq
which describes the evolution of the BM thickness and can be solved using a finite difference approximation. Consider an infinitesimally small volume of BM $dV = 2\pi \cdot R_i(z,t) \cdot h(z,t) \cdot dz$. Then the difference between the flux into and out of the element $dV$ is equal to $\dd (dV)/ \dd t$, which can be written as
\begin{align}
2\pi \frac{\dd}{\dd t} \left( R_i(z,t) \cdot h(z,t) \right) &= \nonumber \\ 
2\pi \cdot R_i(z,t) \cdot h(z,t) & \cdot q(z,t) - 2\pi \cdot R_i(z + dz,t) \cdot h(z + dz,t) \cdot q(z + dz,t),
\end{align}
which is equal to the model equation \eqref{eq:pvs_model}. For a detailed derivation see Supplementary Information 1. The pressure $p(z,t)$ and the displacement of the BM wall $R_i(z,t)$ are, to a very good approximation, determined solely by the arterial pressure (\ie almost independent of the BM flow), see Supplementary Information 2 for more details. $R_i(z,t)$ and $p_z$ can be obtained from a numerical artery model described in the next section.

@@

This work was supported by an EPSRC Doctoral Training Centre grant (EP/G03690X/1). The authors thank Tony Birch for carrying out the middle cerebral artery flow velocity measurements and Roy O. Weller for his continued advice and encouragement.

>>>>

The BM consists of a complex matrix of proteins (extracellular matrix, ECM) that can be interpreted as a porous medium through which soluble metabolites drain via the ISF. We model flow through this medium, in the standard fashion, by Darcy's law in a 2D cylindrical coordinate system. Darcy's law states that the flux $\q$ is related to the fluid pressure gradient $\boldsymbol{\nabla} p$ by
\beq\label{eq:darcy}
  \q = -K(p_z) \boldsymbol{\nabla} p
\eeq
where $K(p_z)$\footnote{Note that the permeability $K$ used here is the ratio of the intrinsic permeability $k$ of the ECM to the viscosity $\mu$ of ISF, that is $K = k / \mu$} is the pressure gradient dependent permeability of the ECM. We allow the permeability $K$ to be a function of the axial component of the ISF pressure gradient $\partial p(z,t) / \partial z = p_z(z,t)$. In order to obtain a net flux up the vessel, counter to the blood flow, we require this function to be increasing in $p_z$, representing the notion that it is harder to push a flow in the positive $z$-direction ($p_z<0$ and $K(p_z)$ small) than in the negative $z$-direction ($p_z>0$ and $K(p_z)$ large).

@@

Because the BM thickness is much smaller than the vessel length it is possible, on using the lubrication approximation, to derive a 1D model (in $z$) for the BM thickness $h(z,t)$; this takes the form
\beq
\frac{\dd}{\dd t} \left( R_i(z,t) \cdot h(z,t) \right) = \frac{\dd}{\dd z} \left( R_i(z,t) \cdot h(z,t) \cdot K(p_{z}) \cdot p_{z} \right),\label{eq:pvs_model}
\eeq
which describes the evolution of the BM thickness and can be solved using a finite difference approximation. Consider an infinitesimally small volume of BM $dV = 2\pi \cdot R_i(z,t) \cdot h(z,t) \cdot dz$. Then the difference between the flux into and out of the element $dV$ is equal to $\dd (dV)/ \dd t$, which can be written as
\begin{align}
2\pi \frac{\dd}{\dd t} \left( R_i(z,t) \cdot h(z,t) \right) &= \nonumber \\ 
2\pi \cdot R_i(z,t) \cdot h(z,t) & \cdot q(z,t) - 2\pi \cdot R_i(z + dz,t) \cdot h(z + dz,t) \cdot q(z + dz,t),
\end{align}
which is equal to the model equation \eqref{eq:pvs_model}. For a detailed derivation see Supplementary Information 1. The pressure $p(z,t)$ and the displacement of the BM wall $R_i(z,t)$ are, to a very good approximation, determined solely by the arterial pressure (\ie almost independent of the BM flow), see Supplementary Information 2 for more details. $R_i(z,t)$ and $p_z$ can be obtained from a numerical artery model described in the next section. To model the protein valve that is being proposed here $K(p_z)$ is defined as a step function
\begin{equation}
  K(p_z) = \begin{cases}
    K_0 & p_z > 0\\
    K_1 & p_z \leq 0
  \end{cases}
\end{equation}
with $K_1 > K_0$.

@@

This work was supported by an EPSRC Doctoral Training Centre grant (EP/G03690X/1) and EPSRC Doctoral Training Partnership grant (EP/N509747/1). The authors thank Tony Birch for carrying out the middle cerebral artery flow velocity measurements and Roy O. Weller for his continued advice and encouragement.
